### 2016-02-12 prepReport/intro.tex
Implemented half of the suggestions of JCF

\section{Motivation}\label{sec:intro_mot}
Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be speeded up in a dedicated hardware structure.

All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to the date generally have between 2-8 cores (mobile devices and general use personal computers),

In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the portin process is not entierely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details. This way, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a supposedly lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance. Furthermore, when processing an incoming stream of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.

\section{Objectives}\label{sec:intro_obj}
Taking into account the considerations presented in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}. Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate the some of the applications portrayed in~\ref{sec:sa_apps}. Lastly, I will work to make this structure reconfigurable on the go, and trying to minimize the reconfiguration time.

@@

\begin{itemize}
    \item \textbf{Supervisor} -- The \href{https://sigarra.up.pt/feup/pt/func\_geral.formview?p\_codigo=210963}{Professor João Canas Ferreira}, auxiliary professor at the Faculty of Engineering of the University of Porto.
    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
\end{itemize}

>>>>

\section{Motivation}\label{sec:intro_mot}
Hitherto, and to the best of my knowledge, all of the published applications that use LSTM are software based, but the parallel nature of the structure hollers for a dedicated hardware realization that can dramatically increase its performance, something that  has only recently been done once~\cite{Chang15} (see Section~\ref{sec:sa_hardware} for further details), and although it improves the processing time when compared to a naïve software solution, it still lacks the ability to perform on-chip learning, and the learning process is performed offline, in a normal CPU, when it also could be sped up by a dedicated hardware structure.

All these techniques are generally implemented in mainstream processors, making use of general high-level or low-level languages, where all the real parallelism is limited to the number of simultaneous threads that we can run on each physical core, which up to now generally have between 2-8 cores (mobile devices and general use personal computers),

In order to parallelize the computations to the fullest extent, a solution is to port it both to a Graphics Processing Unit (GPU) or even a Field-Programmable Gate Array (FPGA), but the porting process is not entierely automatic and to have the least performance drop possible, it has to be explicitly programmed in CUDA/OpenCL for GPUs, and a Hardware Description Language for FPGAs, with an increasing level of complexity and low-level details. Therefore, it is necessary to provide frameworks that can allow an FPGA to quickly reconfigure itself to run these kind of networks on demand, for a particular task that requires them, achieving a lower computation time, and unburdening the CPU from running it, thus saving performance for running other tasks related with the Operating System, for instance. Furthermore, when processing an incoming stream of highly dimensional data, or with high throughput, a CPU solution might not be scalable, and could benefit greatly from a dedicated hardware implementation.

\section{Objectives}\label{sec:intro_obj}
Taking into account the considerations done in~\ref{sec:intro_mot}, I propose to develop a hardware implementation of an LSTM Network, \emph{with on-chip learning}, improving the performance, capability and flexibility of the existing solution of~\cite{Chang15}. Moreover, I also propose to benchmark this solution and try to compare it with the software performance results of~\cite{Tavear13} and, as a secondary objective, try to use my developed structure to replicate some of the applications portrayed in~\ref{sec:sa_apps}. Lastly, I will work to make this structure reconfigurable on the go, and trying to minimize the reconfiguration time.

@@

\begin{itemize}
    \item \textbf{Supervisor} -- The \href{https://sigarra.up.pt/feup/pt/func\_geral.formview?p\_codigo=210963}{Professor João Canas Ferreira}, auxiliary professor at the Faculty of Engineering of the University of Porto.
    \item \textbf{Second Supervisor} -- \href{http://www.cl.cam.ac.uk/~ijpdmt2/}{Ivo Timóteo}, MSc, a Computer Science PhD candidate at Cambridge University, UK, in the field of Artificial Inteligence.
\end{itemize}

### 2016-02-12 prepReport/stateArt.tex
Implemented half of the suggestions of JCF

Over the course of this chapter, I am going to present an overview of the most recent developments related to the work of this thesis, both in terms of existing dedicated hardware implementations~\ref{sec:sa_hardware}, the most relevant work in adapting suitable training algorithms to hardware~\ref{sec:sa_training} and also some of the most relevant applications of LSTM~\ref{sec:sa_apps}, which are not FPGA-based, but demonstrate how LSTM is useful by itself, and how well it competes with other Machine Learning algorithms in terms of long time-series dependences in data.

\section{LSTM Applications (non-FPGA related)}\label{sec:sa_apps} 
LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address the same set of problems where LSTM are employed, namely predicting and producing classification decisions from time-series data \textbf{citation needed!}. A very comprehensive description of applications can be found in one of the initial authors webpage dedicated to the subject~\footnote{\href{http://people.idsia.ch/~juergen/rnn.html}{http://people.idsia.ch/~juergen/rnn.html}}. I will now enumerate some of the bleeding edge applications of LSTM.

\begin{itemize}
    \item \textbf{Handwriting Recognition} -- and LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}. LSTMs have also proven to surpass HMM based models in terms of optical character recognition in printed text, as~\cite{Breuel13} suggests.

@@

    \item \textbf{Handwriting Synthesis} -- a comprehensive study by Graves shows, among other sequence generation tasks such as Text Prediction, the use of LSTM to produce synthetic handwriting, that looks incredibly similar to human-produced handwriting~\cite{Graves13_2}.

@@

    \item \textbf{Biomedical Applications} -- this network architecture was used in a protein homology detection scheme~\cite{Hochreiter07} using the SCOP 1.53 benchmark dataset, displaying a staggering performance when compared to other methods. Similarly, a recent article from 2015 by Sønderby et al. suggested the use of standalone LSTM and also Convolutional LSTM to perform sub cellular localization of proteins, given solely their protein sequence, with an astounding accuracy of 0.902~\cite{Sonderby15}

@@

    \item \textbf{Video and Image Analysis} -- Vinyals et al., at Google, proposed and LSTM network for \textbf{image captioning}, preceded by a Convolutional Neural Network to apply preprocessing to the images~\cite{Vinyals14}. This way, the LSTM works as \emph{sentence generator}, that captions the images, with state of the art performance. 
        Besides image captioning, video captioning is also an interesting topic. Venugopalan et al. recently proposed a CNN + LSTM architecture to translate video sequences to natural language~\cite{Donahue14}, using the Microsoft Research Video Description Corpus as a dataset.There are other similar studies that combine image and video captioning~\cite{Donahue14_2}.

\end{itemize}

\section{Hardware Implementations of LSTM}\label{sec:sa_hardware}
Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently (November 2015) by Chang et al.~\cite{Chang15} in the Computing Research Repository (CoRR). It consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered~ Zedboard implementation. It uses a character-level language model from Andrej Karpathy, written in Lua using the Torch7 framework (the Lua calls are implemented in C, so no performance is lost).

@@

The implementation itself makes an extensive use of the Multiply-and-Accumulate units (MAC) of the FPGA, which since they are limited in number in the target platform, limits the number of neurons that we can deploy in parallel. The authors report a nearly 75\% usage for only 8 LSTM neurons. Although an apparently excessive number, I am left to scrutinize whether the usage of MACs is compulsory to obtain a better performance than a CPU or if it can be discarded, allowing for smaller cells. This way, we can have a layer of more neurons occupying the same area and, hopefully, the same, or even less, resources within the FPGA.

@@

The SPSA algorithm was initially published by Spall in~\cite{Spall98}, and its theoretical details are outlined in Section~\ref{sec:training_lstm}. As of applications of it to the training of general Neural Networks, the earliest examples come from 1995 and 1996 on~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time where the memory resources of digital circuitry were limited, and so most of these structures were analog-based. Its adequacy was also established for \textbf{control problems}, such as those proposed in~\cite{Figueiredo97}, where a Time Delay Neural Network is used to control an unknown plant in a linear feedback system.

In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network in an FPGA (and thus a digital system), achieving promising results in an Altera \textregistered~ FPGA. The article carefully delineates the approach taken, and also the hardware architecture designed, so it is a very good reference for the design that I will have to implement.

>>>>

Over the course of this chapter, I am going to present an overview of the most recent developments related to the work of this thesis, both in terms of existing dedicated hardware implementations~\ref{sec:sa_hardware}, the most relevant work in adapting suitable training algorithms to hardware~\ref{sec:sa_training} and also some of the most relevant applications of LSTM~\ref{sec:sa_apps}, which are not FPGA-based, but demonstrate how LSTM are useful by themselves, and how well it competes with other Machine Learning algorithms in terms of long time-series dependences in data.

\section{LSTM Applications (non-FPGA related)}\label{sec:sa_apps} 
LSTM Networks are nowadays one of the state of the art algorithms in deep-learning, and their performance is superior to that of other kinds of RNNs and Hidden Markov Models, both of which are generally used to address the same set of problems where LSTM are employed, namely predicting and producing classification decisions from time-series data. A very comprehensive description of applications can be found in one of the initial authors webpage dedicated to the subject~\footnote{\href{http://people.idsia.ch/~juergen/rnn.html}{http://people.idsia.ch/~juergen/rnn.html}}. I will now enumerate some of the leading edge applications of LSTM.

\begin{itemize}
    \item \textbf{Handwriting Recognition} -- an LSTM-based network~\cite{Bertolami09}, submitted by a team of the Technische Universität München, won the 2009 ICDAR Handwriting Recognition Contest, achieveing a recognition rate of up to 91\%~\cite{ICDAR09}. LSTMs have also proven to surpass HMM-based models in terms of optical character recognition of printed text, as~\cite{Breuel13} suggests.

@@

    \item \textbf{Handwriting Synthesis} -- a comprehensive study by Graves shows, among other sequence generation tasks such as Text Prediction, that use of LSTM to produce synthetic handwriting, that looks incredibly similar to human-produced handwriting~\cite{Graves13_2}.

@@

    \item \textbf{Biomedical Applications} -- this network architecture was used in a protein homology detection scheme~\cite{Hochreiter07} using the SCOP 1.53 benchmark dataset, displaying a good performance when compared to other methods. Similarly, a recent article from 2015 by Sønderby et al. suggested the use of standalone LSTM and also Convolutional LSTM to perform sub cellular localization of proteins, given solely their protein sequence, with an astounding accuracy of 0.902~\cite{Sonderby15}

@@

    \item \textbf{Video and Image Analysis} -- Vinyals et al., at Google, proposed an LSTM network for \textbf{image captioning}, preceded by a Convolutional Neural Network (CNN) to apply preprocessing to the images~\cite{Vinyals14}. The LSTM works as \emph{sentence generator}, that captions the images with state of the art performance. 
        Besides image captioning, video captioning is also an interesting topic. Venugopalan et al. recently proposed a CNN + LSTM architecture to translate video sequences to natural language~\cite{Donahue14}, using the Microsoft Research Video Description Corpus as a dataset.There are other similar studies that combine image and video captioning~\cite{Donahue14_2}.
\end{itemize}

\section{Hardware Implementations of LSTM}\label{sec:sa_hardware}
Hitherto, there is but one actual implementation of an LSTM network in hardware, published recently (November 2015) by Chang et al.~\cite{Chang15} in the Computing Research Repository (CoRR). It consists on the proposal of an LSTM cell architecture for dedicated hardware, targeting a Xilinx\textregistered~ Zedboard implementation. It uses a character-level language model from Andrej Karpathy~\footnote{\href{https://github.com/karpathy/char-rnn}{https://github.com/karpathy/char-rnn}}, written in Lua using the Torch7 framework~\footnote{\href{http://torch.ch/}{http://torch.ch/}} (the Lua calls are implemented in C, so no performance is lost).

@@

The implementation itself makes an extensive use of the Multiply-and-Accumulate units (MAC) of the FPGA, which since they are limited in number in the target platform, limits the number of neurons that we can deploy in parallel. The authors report a nearly 75\% usage for only 8 LSTM neurons. Although an apparently excessive number, I am left to scrutinize whether the usage of MACs is compulsory to obtain a better performance than a CPU or if it can be discarded, allowing for smaller cells. This way, we can have a layer of more neurons occupying the same area and, hopefully, the same, or even fewer, resources within the FPGA.

@@

The SPSA algorithm was initially published by Spall in~\cite{Spall98}, and its theoretical details are outlined in Section~\ref{sec:training_lstm}. As to its applications to the training of general Neural Networks, the earliest examples come from 1995 and 1996 in~\cite{Maeda95, Cauwen96} where SPSA is used to train a VLSI \emph{Analog} Neural Networks, a time where the memory resources of digital circuitry were limited, and so most of these structures were analog-based. Its adequacy was also established for \textbf{control problems}, such as those proposed in~\cite{Figueiredo97}, where a Time Delay Neural Network is used to control an unknown plant in a linear feedback system.

In 2005, Maeda and Wakamura published a proposed SPSA hardware implementation~\cite{Maeda05} to train an Hopfield Neural Network in an FPGA (and thus a digital system), achieving promising results in an Altera FPGA. The article carefully delineates the approach taken, and also the hardware architecture designed, so it is a very good reference for the design that I will have to implement.
